Polynomial phase estimation by least squares phase unwrapping

### 2013-06-26 paper.tex
added to paper, needs another read, and to go over responses again

\documentclass[draftcls, onecolumn, 11pt]{IEEEtran}
%\documentclass[journal]{IEEEtran}

@@

%\title{Polynomial phase estimation by phase unwrapping 1: Identifiability and strong consistency}
\title{Polynomial phase estimation by phase unwrapping}

@@

Algorithms that approximate a nearest point have also been studied.  One example is Babai's nearest plane algorithm \cite{Babai1986}.  Another pragmatic approximate approach is to employ the sphere decoder, but prune branches in the search tree to ensure that there are at most a finite number, say $K$, remaining branches at any time.  This approach has been studied previously~\cite{Baro_list_sphere_dec_2003,Zhan2006_K_best_sphere_decoder,Mondal_Kbest_again_2009} and resembles what is called the sequential $M$-algorithm~\citep{Anderson1984_seq_coding_alg} from coding theory.  The approaches vary in how branches are pruned.  In our implementation we prune branches using the same greedy metric as that used by the Schnorr and Euchner sphere decoder~\cite{schnorr_euchner_sd_1994,Agrell2002} for deciding the order in which branches are searched.  Following~\cite{Zhan2006_K_best_sphere_decoder} we refer to this as the \emph{$K$-best algorithm}.  We have found that this algorithm produces excellent estimates of polynomial phase coefficients and can be computed in a reasonable amount of time for $N$ less than about 1000.

@@

A proof of the first part of this theorem (strong consistency) is given in Section~\ref{sec:strongconstproof}.  Space restrictions make it impossible prove asymptotic normality in this paper.  A complete proof of normality is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Proofs for the case of $m=0$ were given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and $m=1$ in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proof here takes a similar approach, but requires new techniques.  %Before giving the proof of strong consistency we make the following remarks.

@@

\section{Proof of strong consistency}\label{sec:strongconstproof}
 Substituting \eqref{eq:noise_circ_poly} into $SS$ we obtain
 \begin{align*}
SS\left( \mubf \right) &= \sum_{n=1}^{N}\fracpart{ \fracpart{ \Phi_n + \sum_{k = 0}^{m}{\tilde{\mu}_k n^k} } - \sum_{k = 0}^{m}{\mu_k n^k} }^{2} \\
&= \sum_{n=1}^{N}\fracpart{  \Phi_n + \sum_{k = 0}^{m}{(\tilde{\mu}_k - \mu_k) n^k} }^{2}.
\end{align*}
Let $\lambdabf = \dealias(\tilde{\mubf} - \mubf) = \tilde{\mubf} - \mubf - \pbf$ where $\pbf$ is a lattice point from $L_{m+1}$. From the definition of $L_{m+1}$ we have $p_0 + p_1 n + \dots + p_{m} n^m$ an integer whenever $n$ is an integer, and so
\begin{align*}
\fracpart{\sum_{k=0}^{m}\lambda_k n^k } &= \fracpart{\sum_{k=0}^{m}(\tilde{\mu}_k - \mu_k - p_k) n^k } \\
&= \fracpart{\sum_{k=0}^{m}(\tilde{\mu}_k - \mu_k) n^k }.
\end{align*}
Let
\[
SS\left( \mubf \right) = \sum_{n=1}^{N}\fracpart{  \Phi_n + \sum_{k = 0}^{m}{\lambda_k n^k} }  ^{2} = N S_{N}\left( \lambdabf \right).
 \]
From the definition of the $\dealias(\cdot)$ function $\lambdabf \in B$ so the elements of $\lambdabf$ satisfy
 \begin{equation} \label{eq:identifiability}
 -\frac{0.5}{k!} \leq \lambda_k < \frac{0.5}{k!}.
 \end{equation} 
Now $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$ is the minimiser of $S_{N}$ in $B$.  %We now analyse the minimiser of $S_N$.  
Let
\[
V_N(\lambdabf) =  \expect S_N(\lambdabf) = \frac{1}{N}\sum_{n=1}^{N} \expect \fracpart{  \Phi_{n}+\sum_{k = 0}^{m}{\lambda_k n^k}}  ^{2}.
\]
Using techniques described in~\cite{Pollard_conv_stat_proc_1984,van2009empirical} one can show that
 \begin{equation}\label{eq:SNVNunifmlln}
\sup_{\lambdabf \in B}\sabs{S_N(\lambdabf) - V_N(\lambdabf)} \rightarrow 0  
 \end{equation}
almost surely as $N\rightarrow\infty$.  This type of result is called a~\emph{uniform law of large numbers}~\cite{Pollard_conv_stat_proc_1984,van2009empirical}.  A full proof of~\eqref{eq:SNVNunifmlln} in given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010}.  We now concentrate attention on the minimiser of $V_N$. Because $\Phi_n$ has zero intrinsic mean 
\begin{equation}\label{eq:Efracpartmined}
\expect \fracpart{ \Phi_n + z }^{2}
\end{equation}
is minimised uniquely at $z = 0$ for $z \in [-\nicefrac{1}{2}, \nicefrac{1}{2})$.  Since the intrinsic variance of $\Phi_n$ is $\sigma^2$, when $z = 0$,
\begin{equation}\label{eq:Efracpartphi}
\expect\fracpart{\Phi_1+z}^{2} = \expect\fracpart{\Phi_1}^{2} = \sigma^2,
\end{equation}
and so the minimum attained value is $\sigma^2$.

@@

Figure~\ref{plot:bench} shows the running time of the LSU estimator computed using the $K$-best method versus the HAF estimator for polynomial phase signals of order 3.  The HAF estimator is computed using the fast Fourier transform and has complexity $O(N \log N)$.   The $K$-best method tends to be faster when the SNR is large.  This is due to the algorithm rapidly finding a nearest point, and subsequently terminating branches in its search tree early.  On the other hand, when the SNR is small, all branches in the tree tend to be searched.  Interestingly the LSU estimator computed using the $K$-best algorithm can be \emph{faster} than the HAF when the SNR is large and when $N$ is approximately less than 100.  The LSU estimator can be computed in a reasonable amount of time (a few minutes) when $N=1000$.  The algorithm becomes prohibitively expensive for large $N$.  Benchmarks have been performed with polynomial phase signals of order larger than $3$.  The results are omitted due to space constraints.  The running time of the LSU estimator does not appear to change significantly as $m$ increases.  This is expected since a nearest point in a lattice of approximately the same dimension $N-m-1$ must be computed.  The complexity of the HAF estimator grows linearly with $m$.

>>>>

%\documentclass[draftcls, onecolumn, 11pt]{IEEEtran}
\documentclass[journal]{IEEEtran}

@@

%\title{Polynomial phase estimation by phase unwrapping 1: Identifiability and strong consistency}
\title{Polynomial phase estimation by least squares phase unwrapping}

@@

Algorithms that approximate a nearest point have also been studied.  One example is Babai's nearest plane algorithm \cite{Babai1986}.  Another pragmatic approximate approach is to employ the sphere decoder, but prune branches in the search tree to ensure that there are at most a finite number, say $K$, remaining branches at any time.  This approach has been studied previously~\cite{Baro_list_sphere_dec_2003,Zhan2006_K_best_sphere_decoder,Mondal_Kbest_again_2009} and resembles what is called the sequential $M$-algorithm~\citep{Anderson1984_seq_coding_alg} from coding theory.  The approaches vary in how branches are pruned.  In our implementation we prune branches using the same greedy metric as that used by the Schnorr and Euchner sphere decoder~\cite{schnorr_euchner_sd_1994,Agrell2002} for deciding the order in which branches are searched.  Following~\cite{Zhan2006_K_best_sphere_decoder} we refer to this as the \emph{$K$-best algorithm}.  The $K$-best algorithm requires $O(NK^2 \log K)$ arithmetic operations. The maximum number of branches $K$ is free to be chosen.  For polynomial phase estimation we have found that setting $K=10N$ yields excellent results, and in this case the algorithm requires $O(N^3 \log N)$ operations.  Simulations in Section~\ref{sec:simulations} show that this algorithm produces excellent estimates of polynomial phase coefficients and can be computed in a reasonable amount of time for $N$ less than about 1000.

@@

A proof of the first part of this theorem (strong consistency) is given in Appendix~\ref{sec:strongconstproof}.  Space restrictions make it impossible prove asymptotic normality in this paper.  A complete proof of normality is available in~\cite{McKilliam_arxiv_pps_unwrapping_2012}.  Proofs for the case of $m=0$ were given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010} and $m=1$ in~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009}.  The proof here takes a similar approach, but requires new techniques.  %Before giving the proof of strong consistency we make the following remarks  
A statement similar to Theorem~\ref{thm:asymp_proof} without a complete proof was given in~\cite{McKilliam_polyphase_est_icassp_2011}.  Here, we give a proof.  The results here are also more general than in~\cite{McKilliam_polyphase_est_icassp_2011}, allowing for a wider class of noise distributions.

@@

Figure~\ref{plot:bench} shows the running time of the LSU estimator computed using the $K$-best method versus the HAF estimator for polynomial phase signals of order 3.  The HAF estimator is computed using the fast Fourier transform and has complexity $O(N \log N)$.   As described in Section~\ref{sec:lattice-theory} the LSU estimator computed using the $K$-best method requires in the worst case $O(N^3 \log N)$ operations.  The $K$-best method tends to be faster when the SNR is large.  This is due to the algorithm rapidly finding a nearest point, and subsequently terminating branches in its search tree early.  On the other hand, when the SNR is small, all branches in the tree tend to be searched.  Interestingly the LSU estimator computed using the $K$-best algorithm is comparable with the HAF when the SNR is large and when $N$ is approximately less than 100.  The LSU estimator can be computed in a reasonable amount of time (a few minutes) when $N=1000$.  The algorithm becomes prohibitively expensive for large $N$.  Benchmarks have been performed with polynomial phase signals of order larger than $3$.  The results are omitted due to space constraints.  The running time of the LSU estimator does not appear to change significantly as $m$ increases.  This is expected since a nearest point in a lattice of approximately the same dimension $N-m-1$ must be computed.  The complexity of the HAF estimator grows linearly with $m$.

\appendix

\subsection{Proof of strong consistency}\label{sec:strongconstproof}
 Substituting \eqref{eq:noise_circ_poly} into $SS$ we obtain
 \begin{align*}
SS\left( \mubf \right) &= \sum_{n=1}^{N}\fracpart{ \fracpart{ \Phi_n + \sum_{k = 0}^{m}{\tilde{\mu}_k n^k} } - \sum_{k = 0}^{m}{\mu_k n^k} }^{2} \\
&= \sum_{n=1}^{N}\fracpart{  \Phi_n + \sum_{k = 0}^{m}{(\tilde{\mu}_k - \mu_k) n^k} }^{2}.
\end{align*}
Let $\lambdabf = \dealias(\tilde{\mubf} - \mubf) = \tilde{\mubf} - \mubf - \pbf$ where $\pbf$ is a lattice point from $L_{m+1}$. From the definition of $L_{m+1}$ we have $p_0 + p_1 n + \dots + p_{m} n^m$ an integer whenever $n$ is an integer, and so
\begin{align*}
\fracpart{\sum_{k=0}^{m}\lambda_k n^k } &= \fracpart{\sum_{k=0}^{m}(\tilde{\mu}_k - \mu_k - p_k) n^k } \\
&= \fracpart{\sum_{k=0}^{m}(\tilde{\mu}_k - \mu_k) n^k }.
\end{align*}
Let
\[
SS\left( \mubf \right) = \sum_{n=1}^{N}\fracpart{  \Phi_n + \sum_{k = 0}^{m}{\lambda_k n^k} }  ^{2} = N S_{N}\left( \lambdabf \right).
 \]
From the definition of the $\dealias(\cdot)$ function $\lambdabf \in B$ so the elements of $\lambdabf$ satisfy
 \begin{equation} \label{eq:identifiability}
 -\frac{0.5}{k!} \leq \lambda_k < \frac{0.5}{k!}.
 \end{equation} 
Now $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$ is the minimiser of $S_{N}$ in $B$.  %We now analyse the minimiser of $S_N$.  
Let
\[
V_N(\lambdabf) =  \expect S_N(\lambdabf) = \frac{1}{N}\sum_{n=1}^{N} \expect \fracpart{  \Phi_{n}+\sum_{k = 0}^{m}{\lambda_k n^k}}  ^{2}.
\]
Using techniques described in~\cite{Pollard_conv_stat_proc_1984,van2009empirical} one can show that
 \begin{equation}\label{eq:SNVNunifmlln}
\sup_{\lambdabf \in B}\sabs{S_N(\lambdabf) - V_N(\lambdabf)} \rightarrow 0  
 \end{equation}
almost surely as $N\rightarrow\infty$.  This type of result is called a~\emph{uniform law of large numbers}~\cite{Pollard_conv_stat_proc_1984,van2009empirical}.  A full proof of~\eqref{eq:SNVNunifmlln} in given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010}.  We now concentrate attention on the minimiser of $V_N$. Because $\Phi_n$ has zero intrinsic mean 
\begin{equation}\label{eq:Efracpartmined}
\expect \fracpart{ \Phi_n + z }^{2}
\end{equation}
is minimised uniquely at $z = 0$ for $z \in [-\nicefrac{1}{2}, \nicefrac{1}{2})$.  Since the intrinsic variance of $\Phi_n$ is $\sigma^2$, when $z = 0$,
\begin{equation}\label{eq:Efracpartphi}
\expect\fracpart{\Phi_1+z}^{2} = \expect\fracpart{\Phi_1}^{2} = \sigma^2,
\end{equation}
and so the minimum attained value is $\sigma^2$.

@@
